---
title: "Titanic"
author: "Tshegofatso Kgomo"
date: "2024-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
###Loading Packages
library(haven)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(apaTables)
library(corrgram)
library(corrplot)
library(DescTools)
library(caTools)
library(rpart)
library(rpart.plot) 
library(GGally)
library(reshape2)
library(caret)
```


```{r}
###Loading the Titanic data set
titanic <- read.csv("Titanic-Dataset.csv")
```


```{r}
###Overview of the Titanic data set
summary(titanic)
str(titanic)
head(titanic)
```


```{r}
###Frequency Count
ggplot(titanic, aes(x= Survived))+
  geom_bar()+
  labs(title = "Survivors", X = "Survivors", y= "Count")

ggplot(titanic, aes(x= Survived, fill= Sex))+
  geom_bar(position = "dodge")+
  labs(x= "Survived", y= "Count", fill= "Sex")+
  ggtitle("Survivors by Sex")


ggplot(titanic, aes(x= factor(Survived), fill= factor(Pclass)))+
  geom_bar(position = "dodge")+
  labs(x= "Survived", y= "Count", fill= "Pclass")+
  ggtitle("Survivors by Class")


ggplot(titanic, aes(x= factor(Survived), fill= factor(SibSp)))+
  geom_bar(position = "dodge")+
  labs(x= "Survived", y= "Count", fill= "SibSp")+
  ggtitle("Survivors by Sibling/Spouse")


ggplot(titanic, aes(x= Age))+
  geom_histogram(binwidth = 0.5, fill = "purple", color ="grey")+
  labs(title = "Age", X = "Age", y= "Count")
```

*More people did not survive.*

*More male travellers did not survive.*

*Most of the travellers who did not survive were in 3rd class.*

*Most of the travellers had either no siblings or 1 sibling.*

*Most of the travellers were between 18-36, and the oldest traveller being 80 years*

```{r}
###checking for missing values
titanic[titanic$Name=="", ]
titanic[titanic$Sex=="", ]
titanic[titanic$Cabin=="", ]
sum(is.na(titanic$Age))
```

The Cabin Variable has 687 missing values, and the Age variable has 177 missing values.

```{r}
###Data for analysis
titanic2 <- titanic[, c(2,3,5,6,7,8)]
```

To decision is to drop PassengerID, Name, Ticket, Fare, Cabin, and Embarked. These
are dropped because they will provide much information in this instance of the analysis.


```{r}
###Replacing Missing Values
sum(is.na(titanic2$Age))
md <- Mode(titanic2$Age, na.rm = TRUE)
md
titanic2$Age <- ifelse(is.na(titanic2$Age), md, titanic2$Age)
sum(is.na(titanic2$Age))
```

The missing values in the Age variable are replaced with the mode of the Age variable.

```{r}
###Summary statistics
head(titanic2)
summary(titanic2)
str(titanic2)
```


```{r}
### Using various visuals to see the final data set for analysis
ggplot(titanic2, aes(Survived, Sex))+
  geom_violin(trim = FALSE)+
  ggtitle("survived by sex")

ggplot(titanic2, aes(Survived)) + geom_density()+ ggtitle("Survived")
ggplot(titanic2, aes(Pclass)) + geom_density()+ ggtitle("Pclass")
ggplot(titanic2, aes(Sex)) + geom_density()+ ggtitle("Sex")
ggplot(titanic2, aes(Age)) + geom_density()+ ggtitle("Age")
ggplot(titanic2, aes(SibSp)) + geom_density()+ ggtitle("SibSp")
ggplot(titanic2, aes(Parch)) + geom_density()+ ggtitle("Parch")

data.long <- melt(titanic2)
ggplot(data.long, aes(Sex, fill= variable))+
  geom_density(alpha=0.5)+
  facet_wrap(~variable, scales = "free")+
  theme_minimal()
```



```{r, message=FALSE, results='asis'}
###Correlation between the variables
ggpairs(data = titanic2, columns= c(1,2, 3,4,5,6), title = "Titanic Survivors")
corrgram(titanic2, order = TRUE, lower.panel = panel.shade, upper.panel = panel.pie)
``` 


```{r}
###changing survived variable into factor
new.labels <- c("Died", "Survived")
titanic2$Survived <- factor(titanic2$Survived, labels = new.labels)
table(titanic2$Survived)
titanic2$Sex <- factor(titanic2$Sex)
```


```{r, results='asis', message=FALSE}
###training data
set.seed(101)
sample <- sample.split(titanic2$Survived, SplitRatio = 0.5)

train <- subset(titanic2, sample == TRUE)
test <- subset(titanic2, sample == FALSE)
```
The data is split into training and test data.

```{r}
traindata <- train[,2:6]
trainclass <- train[,1]
tr <- trainControl(method = "cv", number = 5, savePredictions = TRUE)
titanicPredict <- train(traindata, trainclass,
              method = "rpart",
              trControl = tr)

confusionMatrix(titanicPredict)

rpart.plot(titanicPredict$finalModel, box.palette = "BuGn",main = "Survivors", extra = 101)

rpart.rules(titanicPredict$finalModel, cover = TRUE, clip.facs = TRUE)

TestPredict <- rpart.predict(titanicPredict,test)

table(TestPredict, test$Survived) %>%
  prop.table()%>%
  round(3)
```

## Confusion Matrix
The confusion matrix has an accuracy average of 78.8%.

## Decision Tree Model

More male travellers did not survive the titanic crash.

Of those who did not survive, majority were from 3rd class.

Majority of those who survived had less than or 3 sibling and were less than or 30 years of age.

## Predicition
The prediction of the model are not far off, when testing data is used. 







