---
title: "IRIS"
author: "Tshegofatso Kgomo"
date: "2024-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
###Loading Packages
library(haven)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(corrplot)
library(DescTools)
library(caTools)
library(rpart)
library(rpart.plot) 
library(caret)
library(reshape2)
```



```{r}
###Loading the Iris data set
iris <- read.csv("IRIS.csv")
```


```{r}
###Overview of data set
summary(iris)
str(iris)
head(iris)
```
All four variables are numerical vectors, and species is a character vector.

```{r}
###Checking missing values
is.na(iris)
iris$species
iris[iris$species=="", ]
```

The data set does not have any missing values.

```{r}
###Visualizing the data
ggplot(iris, aes(x= sepal_length))+
  geom_bar()+
  labs(title = "Species length", X = "Species length", y= "Count")


ggplot(iris, aes(x= factor(species), fill= factor(sepal_length)))+
  geom_bar(position = "dodge")+
  labs(x= "Species", y= "Count", fill= "sepal_length")+
  ggtitle("Species by length")


ggplot(iris, aes(x= factor(species), fill= factor(sepal_width)))+
  geom_bar(position = "dodge")+
  labs(x= "Species", y= "Count", fill= "sepal_width")+
  ggtitle("Species by width")

ggplot(iris, aes(x= factor(species), fill= factor(petal_length)))+
  geom_bar(position = "dodge")+
  labs(x= "Species", y= "Count", fill= "petal_length")+
  ggtitle("Species by petal length")

ggplot(iris, aes(x= factor(species), fill= factor(petal_width)))+
  geom_bar(position = "dodge")+
  labs(x= "Species", y= "Count", fill= "petal_width")+
  ggtitle("Species by petal width")

data_long <- melt(iris)
ggplot(data_long, aes(species, fill= variable))+
  geom_density(alpha=0.6)+
  facet_wrap(~variable, scales = "free")+
  theme_minimal()
```


```{r}
cor_matrix <- cor(iris[, -5])
corrplot(cor_matrix, method = "color", type = "full", order = "hclust", tl.col="black", tl.srt = 45, addrect = 2)
```


```{r}
###Changing the species into factor
iris$species <- factor(iris$species)
```

Changing the species variable into a factor to make it easier to run through
a predictive model.

```{r}
###Splitting the data
set.seed(101)
sample <- sample.split(iris$species, SplitRatio = 0.5)

Iristest <- subset(iris, sample == FALSE)
Iristrain <- subset(iris, sample == TRUE)
```

Splitting the data into training data, which will be used to train the predictive
model. Then, use the testing data to run the prediction. 

```{r}
###Modeling the data
traindata <- Iristrain[,1:4]
trainclass <- Iristrain[,5]
tr <- trainControl(method = "cv", number = 5, savePredictions = TRUE)
IrisPredict <- train(traindata, trainclass,
              method = "rpart",
              trControl = tr)

confusionMatrix(IrisPredict)

rpart.plot(IrisPredict$finalModel, box.palette = "BuGn",main = "Iris", extra = 101)

rpart.rules(IrisPredict$finalModel, cover = TRUE, clip.facs = TRUE)

TestPredict <- rpart.predict(IrisPredict,Iristest)

table(TestPredict, Iristest$species) %>%
  prop.table()%>%
  round(3)
```

## Confusion Matrix
When the training data was used, the average accuracy of the confusion matrix is 92%.

## Decision Model
Based on the decision tree, the sepal length and width were excluded in the final decision.
Furthermore, the results indicate that: if the petal length is smaller than 2.5 then its Iris-Setosa; If the petal width is smaller than 1.6 then it is Iris-Virginica; and if the petal length is greater than 2.5 and petal width is greater than 1.6 then it is a Iris-veriscolor.

## Predicition
When checking the prediction using test data, the proportion indicate that the prediction are not too far off. 
